Relocating a cluster of earthquakes using a single station

### 2012-07-18 paper2/paper2.tex
another round of changes - We now have a complete draft

\date{Received 2010 Month Day; in original form 2010 Month Day}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{200}
\pubyear{2010}

@@

\begin{summary}
(Type abstract here)
\end{summary}

\begin{keywords}
 Probability distributions; Earthquake source observations; Seismicity and tectonics; Wave scattering and diffraction
\end{keywords}

@@

Earthquake location is important for many applications. It assists
our understanding of seismic activity by indicating regions of
relatively higher seismicity such as lithospheric plate boundaries
\citep[e.g., ][]{dr_Sykes67a, dr_Isacks68a, dr_Stein02a} or local
active source zones \citep[e.g., ][]{dr_Gutenberg44a,
dr_Giardini03a}. Earthquake locations are required for  magnitude
determination \citep[e.g., ][]{dr_Richter35a, dr_Gutenberg45a},
computing moment tensors \citep[e.g., ][]{dr_Sipkin02a} and
seismological studies of the Earth's interior \citep[e.g.,
][]{dr_Spencer80a, dr_Kennett95a, dr_Curtis02a, dr_Kennett04a}. They
are needed to understand strong motion and seismic attenuation
\citep[e.g., ][]{dr_Toro97a, dr_Campbell03a}
 and model earthquake hazard or risk
\citep[e.g., ][]{dr_Frankel00a, dr_Stirling02a, dr_Robinson06b}.

The accuracy required in earthquake location depends on the
application. For example, imaging the structure of a fracture system
from microseismicity requires greater detail than determining
whether a $M_w=7.5$ earthquake occurs offshore for tsunami warning
purposes. This paper focuses on reducing the uncertainty on the
locations of a cluster of events when they are recorded by a small
number of stations.

Absolute location describes the location of an earthquake with
respect to a global reference such as latitude, longitude (or
easting/northing) and depth. Uncertainties associated with absolute
locations are influenced by source to station distances, the number
of stations and their geometry, quality of signal and accuracy of
the velocity model used in computing travel times. Uncertainties in
absolute location are typically of the order of several km because
they are susceptible to uncertainty in the velocity structure along
the entire path between the source and receiver. For example,
\citet{dr_Shearer99a} states that location uncertainty in the ISC
and PDE catalogues are generally around 25\,km horizontally and at
least 25\,km in depth.\footnote{Here the depth uncertainties of
25\,km assume the use of depth dependent phases such as $pP$.
Without such phases the uncertainty is higher.}
  These locations are created using global seismic networks.
\citet{dr_Bondar04a} demonstrate that at the local scale, absolute
locations are accurate to within 5\,km with a 95\% confidence level
when local networks meet a number of station related criteria. Such
errors are too large for many applications, particularly those
focussed on imaging rupture surfaces from aftershock sequences.

Relative earthquake location involves locating a group of
earthquakes with respect to one another and was first introduced by
\citet{dr_Douglas67a} who developed the technique commonly known as
joint hypocenter determination.\footnote{\citet{dr_Douglas67a}
originally used the term joint epicentre determination. However, he
was solving for hypocentre.} In principle, relative locations can be
computed by differencing absolute locations. However,
\citet{dr_Pavlis92a} shows that inadequate knowledge of velocity
structure leads to systematic biases when relative positions are
computed in this way. To reduce errors from unknown velocity
structure, relative location techniques compute locations directly
from travel time differences between two waveforms \citep[e.g.,
][]{dr_Ito85a, dr_Got94a, dr_Nadeau97a, dr_Waldhauser99a}. By doing
so, they remove errors associated with velocity variations outside
the local region, because such variations influence all waveforms in
the same manner \citep{dr_Shearer99a}.

Reported location uncertainties from relative techniques are  around
15 to 75\,m in local settings with good station coverage
\citep[e.g., ][]{dr_Ito85a, dr_Got94a,
dr_Waldhauser99a,dr_Waldhauser08a}. Here, `good coverage' implies
multiple stations distributed across a broad range of azimuthal
directions. Relative location techniques have been applied to a
diverse range of problems including: location of active fault planes
\citep[e.g., ][]{dr_Deichmann92a, dr_Got94a, dr_Waldhauser99a,
dr_Waldhauser02a, dr_Shearer05a}; studying rupture mechanics
\citep[e.g., ][]{dr_Rubin99a, dr_Rubin02a}; interpreting magma
movement in volcanoes \citep[e.g., ][]{dr_Fremont87a}; and
monitoring pumping induced seismicity \citep[e.g., ][]{dr_Lees98a,
dr_Ake05a}. \citet{dr_Poupinet84a}, \citet{dr_Bokelmann00a} and
\citet{dr_Rubin02b} apply relative location techniques to identify
earthquake doublets, events that have near identical locations.

@@

\citet{dr_Snieder05a} demonstrate that the coda of two earthquakes
can be used to estimate the separation between them. Their
technique, known as coda wave interferometry (CWI), is based on the
interference pattern between the coda waves. Unlike travel time
techniques, CWI
 does not require multiple stations or good azimuthal coverage.
In fact, it is possible to obtain estimates of separation using a single
station \citep{dr_Robinson07b}. This makes CWI particularly interesting
for regions where station density is low such as intraplate areas.
Our technique can be used on coda waves alone or in combination with
travel times. We demonstrate that the use of coda waves improves
the location of a cluster of events when the recording situation
is poor and we show that it is possible to obtain accurate relative locations with as
little as one station.

@@

For three earthquakes we have likelihoods;
$P(\widetilde{\delta}_{CWIN,12}|\mathbf{e}_1, \mathbf{e}_2)$,
$P(\widetilde{\delta}_{CWIN,13}|\mathbf{e}_1, \mathbf{e}_3)$
and
$P(\widetilde{\delta}_{CWIN,23}|\mathbf{e}_2, \mathbf{e}_3)$.
In writing
these likelihoods we have replaced the conditional term on separation
$\widetilde{\delta}_t$ with the locations (e.g. $\mathbf{e}_1$ and $\mathbf{e}_2$).
This can be done because
knowledge of location translates to separation.\footnote{Note that the reverse
can not be said. That is, knowledge of separation between a single event pair
does not translate to location but rather places a non-unique constraint on
location.}
Furthermore, since the pairwise functions are independent the joint
likelihood becomes
\begin{equation}
\begin{array}{l}
\label{eq:-joint-likelihood4three}
P(\widetilde{\mathbf{\delta}}_{CWIN} | \mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3) =
P(\widetilde{\delta}_{CWIN,12} | \mathbf{e}_1, \mathbf{e}_2) \\
\hspace{2em} \times P(\widetilde{\delta}_{CWIN,13} | \mathbf{e}_1, \mathbf{e}_3)
\times  P(\widetilde{\delta}_{CWIN,23} | \mathbf{e}_2, \mathbf{e}_3).
\end{array}
\end{equation}
Similarly, we note that earthquake locations are independent and the
joint prior becomes
\begin{equation}
\label{eq:-joint-prior4three}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3) = P((\mathbf{e}_1) \times P(\mathbf{e}_2) \times P(\mathbf{e}_3).
\end{equation}
Combining eqs. (\ref{eq:-joint-likelihood4three}) and (\ref{eq:-joint-prior4three})
gives the joint posterior function
\begin{equation}
\label{eq-3eqposterior}
\begin{array}{l}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \widetilde{\mitbf{\delta}}_{CWIN}) = c \displaystyle \prod_{i=1}^3 P(\mathbf{e}_i) \\
\hspace{4em}  \times \displaystyle \prod_{i=1}^2 \prod_{j=i+1}^3 P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)
\end{array}
\end{equation}
for three events. A detailed understanding of the location of a single event (e.g. $\mathbf{e}_2$) is obtained by computing the
marginal
\begin{equation}
\label{eq:-E2-marginal}
P(\mathbf{e}_2) = \int \int P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \widetilde{\mathbf{\delta}}_{CWIN}) d\mathbf{e}_1 d\mathbf{e}_3
\end{equation}
where the intergral is taken over all plausible locations for $\mathbf{e}_1$ and $\mathbf{e}_3$.
Alternatively, we can compute the marginal for a single event coordinate
by integrating the posterior over all events and remaining coordinates
for the chosen earthquake. Evaluation of $c$ in eq. (\ref{eq-3eqposterior}) involves finding the value which
normalises the posterior function to unit area over all plausible locations.
In many applications the constant of
proportionality $c$ is not important. For example, it is not required
when seeking the combination of locations which maximise
 the posterior function, nor in Bayesian sampling algorithms such as Markov-chain Monte-Carlo
techniques which require evaluation of a function proportional to the PDF.

By following the same argument we get the posterior function
\begin{equation}
\begin{array}{l}
\label{posterior-n-events}
P(\mathbf{e}_1,...,\mathbf{e}_n | \widetilde{\mathbf{\delta}}_{CWIN}) = c \displaystyle \prod_{i=1}^n P(\mathbf{e}_i) \\
\hspace{4em} \displaystyle  \times \prod_{i=1}^{n-1} \prod_{j=i+1}^n P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)
\end{array}
\end{equation}
for $n$ events. When evaluating eq. (\ref{posterior-n-events}) over a range of locations
it is necessary to compute and multiply many numbers close to zero.
This is because the PDFs tend to zero as
the locations get less likely (i.e. near the boundaries of the plausible region).
Such calculations are prone to truncation errors so we
work with the negative logarithm
\begin{equation}
\label{eq:-negative-log-part1}
L(\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n) = -ln\left[ P(\mathbf{e}_1,...,\mathbf{e}_n | \widetilde{\mathbf{\delta}}_{CWIN} )  \right]
\end{equation}
or
\begin{equation}
\begin{array}{l}
\label{eq:-negative-log}
L(\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n) =
-ln\left[ c \right] - \displaystyle \sum_{i=1}^n ln\left[P(\mathbf{e}_i)\right] \\
\hspace{3em} - \displaystyle \sum_{i=1}^{n-1} \sum_{j=i+1}^n ln\left[P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)\right].
\end{array}
\end{equation}
The logarithm improves numerical stability by replacing products with summations. The negative assists with
optimisation, many algorithms for which are designed to minimise an objective function.

The event locations $\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n$
are defined by coordinates $\hat{x}$, $\hat{y}$ and $\hat{z}$
where the hat $\hat{.}$ indicates use of
a local coordinate system. We choose a local coordinate system
which removes ambiguity associated with transformations of the coordinate system,
such as translations and rotations, that do not change the distance between events.
In defining this coordinate system we fix the first event at the origin,
\begin{equation}
\label{eq:coord-const1}
\mathbf{e}_1 = (0,0,0),
\end{equation}
the second event on the positive $\hat{x}$-axis
\begin{equation}
\label{eq:coord-const2}
\mathbf{e}_2 = (\hat{x}_2, 0, 0), \hat{x}_2>0
\end{equation}
the third on the $\hat{x}-\hat{y}$ plane
\begin{equation}
\label{eq:coord-const3}
\mathbf{e}_3 = (\hat{x}_3,\hat{y}_3,0), \hat{y}_3>0
\end{equation}
and the fourth to
\begin{equation}
\label{eq:coord-const4}
\mathbf{e}_4 = (\hat{x}_4,\hat{y}_4,\hat{z}_4), \hat{z}_4>0.
\end{equation}
This coordinate system reduces translational (constraint \ref{eq:coord-const1})
and rotational (constraints \ref{eq:coord-const2} to \ref{eq:coord-const4}) non-uniqueness
without loss of generality. It is necessary to work with a local coordinate
system when using coda waves alone because the CWI technique constrains only event separation
between earthquakes. The inclusion of travel times in Sec. \ref{sec:CalaverasLoc-CWIandTT}
allows us to move to a global reference system.

@@

\subsection{Synthetic experiments}
\label{sec:benchmarking}

@@

\subsection{2D synthetic experiments}

Actual locations are created by
randomly selecting $\hat{x}$- and $\hat{y}$-coordinates such that $-50 \leq \hat{x} \leq 50$ and $-50 \leq \hat{y} \leq 50$.
These are indicated with triangles in Fig. \ref{fig-2D50eq-relocation_eg1}.
 We assume a local velocity of $v=3300\,$ms$^{-1}$ between all
event pairs and a dominant frequency of 2.5$\,$Hz to represent waveform data filtered between 1 and 5$\,$Hz.
The CWI data are defined by the positive bounded Gaussian with statistics
$\bar{\mu}_N$ and $\bar{\sigma}_N$. A hypothetical CWI mean is created by setting
\begin{equation}
\label{eq:hypothetical-CWI-mean-optichapt}
\bar{\mu}_N = \mu_1 \left( \widetilde{\delta}_t \right)
\end{equation}
using eq. (\ref{eq:mu1}). This assumption ensures that the sample
mean of hypothetical separation estimates is consistent with known
CWI biases \citep{dr_Robinson11a}. In example 1 we use
$\bar{\sigma}_N=0.02$ between all event pairs. Application of our
optimisation procedure on the hypothetical CWI data yields the
circles in Fig. \ref{fig-2D50eq-relocation_eg1}. The optimisation
does not lead to the correct solution due to the addition of noise
$\bar{\sigma}_N=0.02$ on the hypothetical CWI data. The average
coordinate error is 2.0\,m which is small compared to the noise of
$\bar{\sigma}_N=0.02$ which for $v_s=3300$\,ms$^{-1}$ and
$f_{dom}=2.5$\,Hz corresponds to roughly 25\,m.
\begin{figure}
\includegraphics[width = 20pc]{diags/locs_2D_50eq_1.eps}
\caption{Example 1 - Synthetic relocation of 50 earthquakes in 2D
using all constraints with noise $\bar{\sigma}_N=0.02$. Actual and
optimisation event locations are shown in triangles and circles,
respectively.} \label{fig-2D50eq-relocation_eg1}
\end{figure}

@@

The local coordinate system works by constraining the location of
the first three earthquakes. Earthquake 1 is fixed at the origin,
earthquake 2 on the positive $\hat{x}$-axis and earthquake 3 has
$\hat{y}>0$. As the number of events increases the strength of this
constraint on later events becomes weaker allowing small rotations
of events with respect to each other. That is, even though the
rotational freedom of the cluster is in principal removed by the
constraints imposed on the events (see eqs. \ref{eq:coord-const1} to
\ref {eq:coord-const4}) we observe that in practice the presence of
noise allows the rotational non-uniqueness to reappear. This is
because the `easiest' way data noise can propagate is into the
direction which is least constrained by the data. The same phenomena
is observed in linear inversion where noise creates large spurious
model changes in directions of the eigenvectors with the smallest
singular values \citep[e.g ][]{dr_Aster05a}. Unfortunately, there is
no obvious way to overcome this issue when using coda waves alone
since the constraints are based on separation. Fortunately however,
combining coda waves with measurements of travel times alleviates
this problem and facilitate the removal of a local coordinate system
altogether.

@@

\subsubsection{The impact of reduced linkage}

@@

In Figs. \ref{fig:optimisationresults-2Dsynth}(a) and (b) we examine
the performance of the optimisation statistically. In particular, we
plot the maximum $\Delta_{max}$ and mean $\Delta_\mu$ of the
coordinate difference as a function of increasing linkage. We show
the statistics for the `best' optimisation solution (black) and for
the solution space when all 25 optimisations are considered (grey).
In this case the best solution is determined by the set of event
locations which lead to the smallest value of $L$. We observe
similar errors in the best solution for all cases when at least 30\%
of the branches are used. The errors increase drastically when only
10\% or 20\% of the constraints are included. Interestingly, this
breakdown around 20\% to 30\% coincides with the point where the
average number of branches required to link each pair reaches 2 (see
Fig.\ref{fig:optimisationresults-2Dsynth}(c)). Since the average
number of branches can be computed in advance it can be used as an
indication of the potential stability of the inversion prior to
optimisation. An earlier breakdown is observed when all 25 solutions
are considered collectively. For example, the maximum coordinate
error $\Delta_{max}$ is significantly higher than that for the best
solution for all cases with 60\% of the constraints or less. This
observation confirms that the optimisation is susceptible to the
presence of local minima and that a range of starting points should
be considered.

Interestingly, we note that some of the optimisations chains
fail to converge after 1200 iterations when the linkage is 60\% or lower and all
fail when then linkage is 20\% or lower. Despite their failure to converge however, many of the
solutions remain relatively close to the actual solutions. The derivatives used in the conjugate gradient method
depend on events connected by CWI measurements. Consequently,
earthquakes that are only connected via other events do not `communicate' with each other
directly. To some extent, this should be taken care of during the iterative process where location information
can spread to events which have no direct links. However, the
lack of connection through the gradient could prevent convergence in extreme cases, or more likely
slow the procedure down. This could explain why some
examples do not converge after 1200 iterations.
\citet{dr_VanDecar94a} show that gradient damping acts
slowly through iterative least-squares, because
 every cell in one iteration communicates only with its neighbours, and they demonstrate that this can be
fixed with preconditioning in some cases. Their findings suggest that it may be possible to improve
the convergence (stability and/or speed) of the CWI optimisation by preconditioning.

\subsubsection{3D synthetic examples with reduced linkage}

>>>>

\date{Received 2012 Month Day; in original form 2012 Month Day}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{200}
\pubyear{2012}

@@

\begin{summary}
 Coda wave interferometry (CWI) is an emerging tool for constraining 
earthquake source properties from the intereference pattern of coda waves between nearby
events. A new earthquake location algorithm is derived which relies on CWI based probabilistic estimates 
of earthquake separation. The algorithm can be used with CWI estimates alone or in tandem with travel time
data. Synthetic examples in 2D and 3D and real earthquakes on the 
Calaveras Fault, California are used to demonstrate the usefulness of the CWI technique in locating 
poorly recorded earthquakes. It is demonstrated that CWI: (a) outperforms traditional earthquake location 
techniques when the number of stations is small; (b) is self-consistent across a broad range of station
situations; and (c) can be used with a single station to locate earthquakes.

\end{summary}

%\begin{keywords}
% Probability distributions; Earthquake source observations; 
%Seismicity and tectonics; Wave scattering and diffraction
%\end{keywords}

@@

Earthquake location is important for many applications. Locations 
are required for:  magnitude
determination \citep{dr_Richter35a, dr_Gutenberg45a};
computing moment tensors \citep{dr_Sipkin02a};
seismological studies of the Earth's interior 
\citep{dr_Spencer80a, dr_Kennett95a, dr_Curtis02a, dr_Kennett04a};
understanding strong motion and seismic attenuation
\citep{dr_Toro97a, dr_Campbell03a}
 and modeling earthquake hazard or risk
\citep{dr_Frankel00a, dr_Stirling02a, dr_Robinson06b}.

The accuracy required in earthquake location depends on the
application. For example, imaging the structure of a fracture system
from microseismicity requires greater detail than determining
whether a $M_w=7.5$ earthquake occurs offshore for tsunami warning.
 This paper focuses on reducing location uncertainty for 
a cluster of events when they are recorded by a small
number of stations.

Absolute location describes the location of an earthquake with
respect to a global reference such as latitude, longitude (or
easting/northing) and depth. Uncertainties associated with absolute
locations are influenced by source to station distances, the number
of stations and their geometry, quality of signal and accuracy of
the velocity model used in computing travel times. Uncertainties in
absolute location are typically of the order of several km because
they are susceptible to uncertainty in the velocity structure along
the entire path between the source and receiver. For example,
\citet{dr_Shearer99a} states that location uncertainty in the ISC
and PDE catalogues are generally around 25\,km horizontally and at
least 25\,km in depth.\footnote{Here the depth uncertainties of
25\,km assume the use of depth dependent phases such as $pP$.
Without such phases the uncertainty is higher.}
\citet{dr_Bondar04a} demonstrate that at the local scale, absolute
locations are accurate to within 5\,km with a 95\% confidence level
when local networks meet a number of station related criteria. Such
errors are too large for many applications, particularly those
focussed on imaging rupture surfaces from aftershock sequences.

Relative earthquake location involves locating a group of
earthquakes with respect to one another and was first introduced by
\citet{dr_Douglas67a} who developed the technique commonly known as
joint hypocenter determination.\footnote{\citet{dr_Douglas67a}
originally used the term joint epicentre determination. However, he
was solving for hypocentre.} In principle, relative locations can be
computed by differencing absolute locations. However,
\citet{dr_Pavlis92a} shows that inadequate knowledge of velocity
structure leads to systematic biases when relative positions are
computed in this way. To reduce errors from unknown velocity
structure, relative location techniques compute locations directly
from travel time differences between two waveforms 
\citep{dr_Ito85a, dr_Got94a, dr_Nadeau97a, dr_Waldhauser99a}. By doing
so, they remove errors associated with velocity variations outside
the local region, because such variations influence all waveforms in
the same manner \citep{dr_Shearer99a}.

Reported location uncertainties from relative techniques are  around
15 to 75\,m in local settings with good station coverage
\citep{dr_Ito85a, dr_Got94a,
dr_Waldhauser99a,dr_Waldhauser08a}. Here, `good coverage' implies
multiple stations distributed across a broad range of azimuthal
directions. Relative location techniques have been applied to a
diverse range of problems including: location of active fault planes
\citep{dr_Deichmann92a, dr_Got94a, dr_Waldhauser99a,
dr_Waldhauser02a, dr_Shearer05a}; studying rupture mechanics
\citep{dr_Rubin99a, dr_Rubin02a}; interpreting magma
movement in volcanoes \citep{dr_Fremont87a}; and
monitoring pumping induced seismicity \citep{dr_Lees98a,
dr_Ake05a}.

@@

\citet{dr_Snieder05a} demonstrate that the coda of two earthquakes
can be used to estimate the separation between them. Their
technique, known as coda wave interferometry (CWI), is based on the
interference pattern between the coda waves. Unlike travel time
techniques, CWI
 does not require multiple stations or good azimuthal coverage.
In fact, it is possible to obtain estimates of separation using a single
station \citep{dr_Robinson07b}. This makes CWI particularly interesting
for regions where station density is low such as intraplate settings. In this 
paper we demonstrate how CWI separation estimates can be used to constrain 
location with data from a single station. 
Our technique can be used on coda waves alone or in combination with
travel times.

@@

For three earthquakes we have likelihoods;
$P(\widetilde{\delta}_{CWIN,12}|\mathbf{e}_1, \mathbf{e}_2)$,
$P(\widetilde{\delta}_{CWIN,13}|\mathbf{e}_1, \mathbf{e}_3)$
and
$P(\widetilde{\delta}_{CWIN,23}|\mathbf{e}_2, \mathbf{e}_3)$.
In writing
these likelihoods we have replaced the conditional term on separation
$\widetilde{\delta}_t$ with the locations (e.g. $\mathbf{e}_1$ and $\mathbf{e}_2$).
This can be done because
knowledge of location translates to separation.\footnote{Note that the reverse
can not be said. That is, knowledge of separation between a single event pair
does not translate to location but rather places a non-unique constraint on
location.}
Furthermore, since the pairwise functions are independent the joint
likelihood becomes
\begin{equation}
\begin{array}{l}
\label{eq:-joint-likelihood4three}
P(\widetilde{\mathbf{\delta}}_{CWIN} | \mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3) =
P(\widetilde{\delta}_{CWIN,12} | \mathbf{e}_1, \mathbf{e}_2) \\
\hspace{2em} \times P(\widetilde{\delta}_{CWIN,13} | \mathbf{e}_1, \mathbf{e}_3)
\times  P(\widetilde{\delta}_{CWIN,23} | \mathbf{e}_2, \mathbf{e}_3).
\end{array}
\end{equation}
Similarly, the earthquake locations are independent and the
joint prior becomes
\begin{equation}
\label{eq:-joint-prior4three}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3) = P((\mathbf{e}_1) \times P(\mathbf{e}_2) \times P(\mathbf{e}_3).
\end{equation}
Combining eqs. (\ref{eq:-joint-likelihood4three}) and (\ref{eq:-joint-prior4three})
gives the joint posterior function
\begin{equation}
\label{eq-3eqposterior}
\begin{array}{l}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \widetilde{\mitbf{\delta}}_{CWIN}) = c \displaystyle \prod_{i=1}^3 P(\mathbf{e}_i) \\
\hspace{4em}  \times \displaystyle \prod_{i=1}^2 \prod_{j=i+1}^3 P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)
\end{array}
\end{equation}
for three events.

A detailed understanding of the location of a single event (e.g. $\mathbf{e}_2$) is obtained by computing the
marginal
\begin{equation}
\label{eq:-E2-marginal}
P(\mathbf{e}_2) = \int \int P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \widetilde{\mathbf{\delta}}_{CWIN}) d\mathbf{e}_1 d\mathbf{e}_3
\end{equation}
where the intergral is taken over all plausible locations for $\mathbf{e}_1$ and $\mathbf{e}_3$.
Alternatively, we can compute the marginal for a single event coordinate
by integrating the posterior over all events and remaining coordinates
for the chosen earthquake. Evaluation of $c$ in eq. (\ref{eq-3eqposterior}) involves finding the value which
normalises the posterior function to unit area over all plausible locations.
In many applications the constant of
proportionality $c$ is not important. For example, it is not required
when seeking the combination of locations which maximise
 the posterior function, nor in Bayesian sampling algorithms such as Markov-chain Monte-Carlo
techniques which require evaluation of a function proportional to the PDF.

Following the same argument we get the posterior function
\begin{equation}
\begin{array}{l}
\label{posterior-n-events}
P(\mathbf{e}_1,...,\mathbf{e}_n | \widetilde{\mathbf{\delta}}_{CWIN}) = c \displaystyle \prod_{i=1}^n P(\mathbf{e}_i) \\
\hspace{4em} \displaystyle  \times \prod_{i=1}^{n-1} \prod_{j=i+1}^n P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)
\end{array}
\end{equation}
for $n$ events. When evaluating eq. (\ref{posterior-n-events}) over a range of locations
it is necessary to compute and multiply many numbers close to zero.
This is because the PDFs tend to zero as
the locations get less likely (i.e. near the boundaries of the plausible region).
Such calculations are prone to truncation errors so we
work with the negative logarithm
\begin{equation}
\label{eq:-negative-log-part1}
L(\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n) = -ln\left[ P(\mathbf{e}_1,...,\mathbf{e}_n | \widetilde{\mathbf{\delta}}_{CWIN} )  \right]
\end{equation}
or
\begin{equation}
\begin{array}{l}
\label{eq:-negative-log}
L(\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n) =
-ln\left[ c \right] - \displaystyle \sum_{i=1}^n ln\left[P(\mathbf{e}_i)\right] \\
\hspace{3em} - \displaystyle \sum_{i=1}^{n-1} \sum_{j=i+1}^n ln\left[P(\widetilde{\delta}_{CWIN,ij}|\mathbf{e}_i,\mathbf{e}_j)\right].
\end{array}
\end{equation}
The logarithm improves numerical stability by replacing products with summations. The negative assists with
optimisation, many algorithms for which are designed to minimise an objective function.

The event locations $\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n$
are defined by coordinates $\hat{x}$, $\hat{y}$ and $\hat{z}$
where the hat $\hat{.}$ indicates use of
a local coordinate system. We choose a local coordinate system
which removes ambiguity associated with transformations of the coordinate system.
In defining this coordinate system we fix the first event at the origin,
\begin{equation}
\label{eq:coord-const1}
\mathbf{e}_1 = (0,0,0),
\end{equation}
the second event on the positive $\hat{x}$-axis
\begin{equation}
\label{eq:coord-const2}
\mathbf{e}_2 = (\hat{x}_2, 0, 0), \hat{x}_2>0
\end{equation}
the third on the $\hat{x}-\hat{y}$ plane
\begin{equation}
\label{eq:coord-const3}
\mathbf{e}_3 = (\hat{x}_3,\hat{y}_3,0), \hat{y}_3>0
\end{equation}
and the fourth to
\begin{equation}
\label{eq:coord-const4}
\mathbf{e}_4 = (\hat{x}_4,\hat{y}_4,\hat{z}_4), \hat{z}_4>0.
\end{equation}
This coordinate system reduces translational (constraint \ref{eq:coord-const1})
and rotational (constraints \ref{eq:coord-const2} to \ref{eq:coord-const4}) non-uniqueness
without loss of generality. It is necessary to work with a local coordinate
system when using coda waves alone because the CWI technique constrains only event separation
between earthquakes. The inclusion of travel times in Section \ref{sec:CalaverasLoc-CWIandTT}
allows us to move to a global reference system.

@@

\section{Synthetic experiments}
\label{sec:benchmarking}

@@

\subsection{Examples 1 and 2 - 2D synthetic experiments}

We design a 2D synthetic experiment (Example 1) by
randomly selecting $\hat{x}$- and $\hat{y}$-coordinates such that $-50 \leq \hat{x} \leq 50$ and $-50 \leq \hat{y} \leq 50$.
These are indicated with triangles in Fig. \ref{fig-2D50eq-relocation_eg1}.
 We assume a local velocity of $v=3300\,$ms$^{-1}$ between all
event pairs and a dominant frequency of 2.5$\,$Hz to represent waveform data filtered between 1 and 5$\,$Hz.
The CWI data are defined by the positive bounded Gaussian with statistics
$\bar{\mu}_N$ and $\bar{\sigma}_N$. A hypothetical CWI mean is created by setting
\begin{equation}
\label{eq:hypothetical-CWI-mean-optichapt}
\bar{\mu}_N = \mu_1 \left( \widetilde{\delta}_t \right)
\end{equation}
using eq. (\ref{eq:mu1}). This assumption ensures that the sample
mean of hypothetical separation estimates is consistent with known
CWI biases \citep{dr_Robinson11a}. In example 1 we use
$\bar{\sigma}_N=0.02$ between all event pairs. Application of our
optimisation procedure on the hypothetical CWI data yields the
circles in Fig. \ref{fig-2D50eq-relocation_eg1}. The optimisation
does not lead to the correct solution due to the addition of noise
($\bar{\sigma}_N=0.02$) on the hypothetical CWI data. The average
coordinate error is 2.0\,m which is small compared to the noise of
$\bar{\sigma}_N=0.02$ which for $v_s=3300$\,ms$^{-1}$ and
$f_{dom}=2.5$\,Hz corresponds to roughly 25\,m.
\begin{figure}
\includegraphics[width = 20pc]{diags/locs_2D_50eq_1.eps}
\caption{Example 1 - Synthetic relocation of 50 earthquakes in 2D
using all constraints with noise $\bar{\sigma}_N=0.02$. Actual and
optimisation event locations are shown in triangles and circles,
respectively.} \label{fig-2D50eq-relocation_eg1}
\end{figure}

@@

The local coordinate system works by constraining the location of
the first three earthquakes. Earthquake 1 is fixed at the origin,
earthquake 2 on the positive $\hat{x}$-axis and earthquake 3 has
$\hat{y}>0$. As the number of events increase the strength of these
constraints on later events weakens allowing small rotations
of events with respect to each other. That is, even though the
rotational freedom of the cluster is in principal removed by the
constraints imposed on the events (see equations (\ref{eq:coord-const1}) to
(\ref{eq:coord-const4})) we observe that in practice the presence of
noise allows the rotational non-uniqueness to reappear. This is
because the `easiest' way data noise can propagate is into the
direction which is least constrained by the data. The same phenomena
is observed in linear inversion where noise creates large spurious
model changes in directions of the eigenvectors with the smallest
singular values \citep{dr_Aster05a}. Unfortunately, there is
no obvious way to overcome this issue when using coda waves alone
since the constraints are based on separation. Fortunately however,
combining coda waves with measurements of travel times alleviates
this problem and facilitate the removal of a local coordinate system
altogether.

@@

\subsection{Example 3 - The impact of reduced linkage}

@@

Figure \ref{fig:optimisationresults-2Dsynth}(a) and (b) illustrates 
the maximum $\Delta_{max}$ (top) and mean $\Delta_\mu$ (middle) of the
coordinate difference as a function of increasing linkage. We show
the statistics for the `best' optimisation solution (black) and for
the solution space when all 25 optimisations are considered (grey).
In this case the best solution is determined by the set of event
locations which lead to the smallest value of $L$. The
error in the best solution is consistent when 30\%
or more of the branches are used. The errors increase when only
10\% or 20\% of the constraints are included. Interestingly, this
breakdown around 20\% to 30\% coincides with the point where the
average number of branches required to link each pair reaches 2 (see
Fig.\ref{fig:optimisationresults-2Dsynth} (bottom)). Since the average
number of branches can be computed in advance it can be used as an
indication of the inversion stability prior to
optimisation. An earlier breakdown is observed when all 25 solutions
are considered collectively. For example, the maximum coordinate
error $\Delta_{max}$ exceeds that for the best
solution for linkage $\leq$60\% confirming that the optimisation is susceptible to
local minima and that a range of starting points should
be considered.

Some optimisations 
fail to converge after 1200 iterations when the linkage is 60\% or lower. All optimisations
fail when then linkage is 20\% or lower. Despite their failure to converge however, the locations
at final iteration are close to the actual solution. The derivatives used in the conjugate gradient method
depend on events connected by CWI measurements. Consequently,
earthquakes that are only connected via other events do not `communicate' with each other
directly. To some extent, this should be adressed during the iterative process where location information
can spread to events which have no direct links. However, the
lack of direct connection through the gradient could prevent convergence in extreme cases, or more likely
slow the procedure down. This could explain why some
examples do not converge after 1200 iterations.
\citet{dr_VanDecar94a} show that gradient damping acts
slowly through iterative least-squares, because
 every cell in one iteration communicates only with its neighbours, and they demonstrate that this can be
fixed with preconditioning in some cases. Their findings suggest that it may be possible to improve
the convergence (stability and/or speed) of the CWI optimisation by preconditioning.

\subsection{Example 4 - 3D synthetic examples with reduced linkage}

In Example 4 we expand the optimisation routine to 3D by randomly picking
 a set of actual event locations for 50 earthquakes with
$-50$\,m$\leq \hat{x},\hat{y},\hat{z} \leq 50$\,m. As in the 2D case we assume a local velocity
of $v=3300\,$ms$^{-1}$ between all event pairs and a dominant frequency of 2.5$\,$Hz to represent
 waveform data filtered between 1 and 5$\,$Hz.
The hypothetical CWI mean is created using equation (\ref{eq:hypothetical-CWI-mean-optichapt})
which ensures consistency between the sample mean of hypothetical separation estimates and CWI
biases. We use a standard deviation for the noisy CWI estimates
of $\bar{\sigma}_N = \sigma_{1to5Hz}$ and perform the optimisation using 10\%,
20\%, ..., 100\% of the direct links.
In each case we repeat the
optimisation 25 times using randomly chosen starting locations. The results are
summarised in Figure \ref{fig:optimisationresults-3Dsynth}.
